# Introduction to Monte Carlo

Today, we'll start the first main topic of the module, which is called "Monte Carlo estimation".

::: {.hidden}
$$\newcommand{\Exg}{\operatorname{\mathbb{E}}} 
\newcommand{\Ex}{\mathbb{E}} $$
:::

## What is statistical computing?

"Statistical computing" -- or "computational statistics" -- as a subject refers to the branch of statistics that involves not attacking statistical problems merely with a pencil and paper, but rather by combining human ingenuity with the immense calculating powers of computers.

One of the big ideas here is **simulation**. Simulation is the idea that we can understand the properties of a random model not by cleverly working out the properties from scratch -- this is usually impossible for anything but the simplest "toy models" -- but rather by running the model many times on a computer. From these many simulations, we can observe and measure things like the typical (or "average") behaviour, the spread (or "variance") of the behaviour, and many other things. This concept of simulation is at the heart of the module MATH5835M Statistical Computing.

In particular, we will look at **Monte Carlo** estimation. Monte Carlo is about estimating a parameter, expectation or probability of a random variable by taking many samples of that random variable, then computing a relevant sample mean from those samples. We will study Monte Carlo in its standard "basic" form (Lectures 2--9), but also in the modern **Markov chain Monte Carlo** form (Lectures 17--23), which has become such a crucial part of Bayesian statistical analysis.

To run a simulation, one needs random numbers with the correct distribution. **Random number generation** (Lectures 10--16) will be an important part of this module. We will look first at how to generate randomness of any sort, and then how to get that randomness into the shape of the distributions we want.

When dealing with a very big data set, traditionally we want to "reduce the dimension" by representing it with a simple parametric model. For example, tens of thousands of datapoints might get reduced just to estimates of the parameters $\mu$ and $\sigma^2$. But with computational statistics, we don't need to make this simplification -- we can do inference using the full details of the whole dataset. An computational idea that takes advantage of this is the **bootstrap** (Lectures 24--27)

## What is Monte Carlo estimation?

Let $X$ be a random variable. We recall the **expectation** $\Ex X$ of $X$: if $X$ is discrete with probability mass function (PMF)  $p$, then this is
$$ \Ex X = \sum_x x\,p(x) ;$$
while if $X$ is continuous with probability density function (PDF) $f$, then this is
$$ \Ex X = \int_{-\infty}^{+\infty} x\,f(x)\,\mathrm{d}x . $$
More generally, the expectation of a function $\phi$ of $X$ is
$$ \Exg \phi(X) = \begin{cases} {\displaystyle \sum_x \phi(x)\,p(x)} & \text{for $X$ discrete}\\ {\displaystyle \int_{-\infty}^{+\infty} \phi(x)\,f(x)\,\mathrm{d}x}  & \text{for $X$ continuous.} \end{cases}$$
(This matches with the "plain" expectation when $\phi(x) = x$.)

But how do we actually *calculate* an expectation like one of these? If $X$ is discrete and can only take a small, finite number of values, we can simply add up the sum $\sum_x \phi(x)\,p(x)$. Otherwise, we just have to hope that $\phi$ and $p$ or $f$ are sufficiently "nice" that we can manage to work out the sum/integral using a pencil and paper. But while this is often the case in the sort of "toy example" one comes across in maths or statistics lectures, this is very rare in "real life".

**Monte Carlo estimation** is the idea that we can get an approximate answer for $\Ex X$ or $\Exg \phi(X)$ if we have access to lots of samples from $X$. For example, if we have access to $X_1, X_2 \dots, X_n$ , independent and identically distributed (**IID**) samples with the same distribution as $X$, then we already know that the mean
$$ \overline X = \frac{1}{n}(X_1 + X_2 + \cdots + X_n) = \frac{1}{n} \sum_{i=1}^n X_i $$
is usually close to the expectation $\Ex X$, at least if $n$ is big. Similarly, it should be the case that
$$ \frac{1}{n} \big(\phi(X_1) + \phi(X_2) + \cdots + \phi(X_n) \big) = \frac{1}{n} \sum_{i=1}^n \phi(X_i) $$
should be close to $\Exg \phi(X)$.

In this course we will write that $X_1, X_2, \dots, X_n$ is a "**random sample** from $X$" to mean that $X_1, X_2, \dots, X_n$ are IID with the same distribution as $X$.

::: {#def-MCest}
Let $X$ be a random variable, $\phi$ a function, and write $\theta = \Exg\phi(X)$. Suppose that $X_1, X_2, \dots, X_n$ are a random sample from $X$. Then the **Monte Carlo estimate** $\widehat\theta_n^{\mathrm{MC}}$ of $\theta$ is
$$ \widehat{\theta}_n^{\mathrm{MC}} = \frac{1}{n} \sum_{i=1}^n \phi(X_i) . $$
:::

While general ideas for estimating using simulation go back a long time, the modern theory of Monte Carlo estimation was developed by the physicists [Stanislaw Ulam](https://en.wikipedia.org/wiki/Stanisław_Ulam) and [John von Neumann](John von Neumann). Ulam (who was Polish) and von Neumann (who was Hungarian) moved to the US in the early 1940s to work on the Manhattan project to build the atomic bomb (as made famous by the film Oppenheimer). Later in the 1940s, they worked together in the Los Alamos National Laboratory continuing their research on nuclear weapons, where they used simulations on early computers to help them numerically solve difficult mathematical and physical problems.

The name "Monte Carlo" was chosen because the use of randomness to solve such problems reminded them of gamblers in the casinos of Monte Carlo, Monaco. Ulam and von Neumann also worked closely with another colleague Nicholas Metropolis, whose work we will study later in this module.

## Examples

Let's see some simple examples of Monte Carlo estimation using R.

::: {#exm-MCexp}
Let's suppose we've forgotten the expectation of the exponential distribution $X \sim \operatorname{Exp}(2)$ with rate 2. In this simple case, we could work out the answer using the PMF $f(x) = 2\mathrm{e}^{2x}$ as
$$ \Ex X = \int_0^\infty x\,2\mathrm{e}^{2x}\,\mathrm{d}x $$
(and, without too much difficulty, get the answer $\frac12$). But instead, let's do this the Monte Carlo way.

In R, we can use the `rexp()` function to get IID samples from the exponential distribution: the full syntax is `rexp(n, rate)`, which gives `n` samples from an exponential distribution with rate `rate`. So our code here should be

```{r, cache = TRUE}
n <- 100
samples <- rexp(n, 2)
MCest <- (1 / n) * sum(samples)
MCest
```

So our Monte Carlo estimate is `r format(round(MCest, 5), nsmall = 5)`, to 5 decimal places.

To get a (probably) more accurate estimation, we could use more samples. We could also simplify the third line of this code by using the `mean()` function.

```{r, cache = TRUE}
n <- 1e6
samples <- rexp(n, 2)
MCest <- mean(samples)
MCest
```

(In the second line, `1e6` is R code for the scientific notation $1 \times 10^6$, or a million.)

Our new Monte Carlo estimate is `r format(round(MCest, 5), nsmall = 5)`, which is much closer to the true value of $\frac12$.
:::

By the way: all R code "chunks" displayed in the notes should work perfectly if you copy-and-paste them into RStudio. I strongly encourage playing about with the code as a good way to learn this material and explore further!

::: {#exm-MC2}
Let's try another example. Let $X \sim \operatorname{N}(1, 2^2)$ be a normal distribution with mean 0 and standard deviation 2. Suppose we want to find out $\Exg(\sin X)$ (for some reason). While it *might* be possible to somehow calculate the integral
$$ \Exg(\sin X) = \int_{-\infty}^{+\infty} (\sin x) \, \frac{1}{\sqrt{2\pi\times 2^2}} \exp\left(\frac{(x - 1)^2}{2\times 2^2}\right) \, \mathrm{d} x , $$
that looks extremely difficult to me.

Instead, a Monte Carlo estimation of $\Exg(\sin X)$ is very straightforward.

```{r}
n <- 1e6
samples <- rnorm(n, 1, 2)
MCest <- mean(sin(samples))
MCest
```
  
Our Monte Carlo estimate is `r format(round(MCest, 5), nsmall = 5)`.
:::

<!--
## Random samples in R

In this lecture, we've got random samples from distributions using the R functions `rexp()` (exponential distribution) and `rnorm()` (normal).

Generally, R has built in functions for generating random samples for the most common distributions. These include:

* `rbinom(n, size, prob)` for `n` samples from the binomial distribution $\operatorname{Bin}(\mathtt{size}, \texttt{prob})$. Includes `rbinom(n, 1, prob)` for the Bernoulli distribution $\operatorname{Bern}(\mathtt{prob})$.

* `rgeom(n, prob)` for `n` samples from the geometric distribution $\operatorname{Geom}(\mathtt{prob})$. (R uses the "number of failures *before* for the first success" definition of the geometric, starting from 0.)

* `rpois(n, lambda)` for `n` samples from the Poisson distribution $\operatorname{Po}(\mathtt{lambda})$.

* `rexp(n, rate)` for `n` samples from the exponential distribution $\operatorname{Exp}(\mathtt{rate})$. If the `rate` parameter is omitted, then `rate = 1` is assumed by default.

* `rnorm(n, mean, sd)` for `n` samples from the normal distribution $\operatorname{N}(\mathtt{mean},\mathtt{sd}^2)$. The third argument is the standard deviation $\sigma$, not the variance $\sigma^2$. If the `mean` and `sd` parameters are omitted, then `mean = 0` and `sd = 1` are assumed by default.

* `runif(n, min, max)` for `n` samples from the continuous uniform distribution on the interval $[\mathtt{min}, \mathtt{max}]$. If the `min` and `max` parameters are omitted, the interval $[0,1]$ is assumed by default. (This will turn out to be the most important sampling distribution in this module.)

* Look at the help file `?distributions` in R for details of other distributions.

Also, we can sample from a discrete random variable on a finite number of outcomes with
```r
sample(x, n, replace = TRUE, prob = p)
```
Here, `x` is a vector of values the random variable can take, `n` is the number of samples, and `p` is a vector of probabilities of each value of `x` (in the same order). If `prob = p` is omitted, equal uniform probabilities of each outcome are assumed by default.

In Part II of this module, we will learn how these R functions work, so we understand *how* R manages to sample from these distributions. Then we will be able to do it ourselves "from scratch", and also be able to sample from distributions that R doesn't have built-in functions for.

-->

**Next time:** *We look at more examples of things we can estimate using the Monte Carlo method.*
