# Envelope rejection sampling I

Last time, we started looking at rejection sampling. We found that if
you start with a proposal distribution $Y$ with PDF $g$, then accept a
proposal $Y = x$ with probability $\alpha(x)$, then the PDF $f$ of a
sample conditional on acceptance satisfies
$f(x) \propto \alpha(x) g(x)$.

The problem is that this sort of gets the problem backwards. If we pick
a proposal distribution $g$ and acceptance function $\alpha$, this tells
us what the sample distribution $f$ will turn out to be. But we are
interested in the reverse question: If we want to target a distribution
$f$, what proposal distribution $g$ and acceptance function $\alpha$ do
we need to get it?

Today we look at how to do that with the "envelope" method.

## Sampling under the curve

We will start with a slightly different discussion, to try to motivate
our approach. (If at any point you find this section confuses more than
it helps, you can skip straight to the definition in the next section.)

Consider a PDF $f$, and draw $f$ as a curve. We know that $f(x)$ is
always positive, so this curve is always above the x-axis. We also know
that $\int_{-\infty}^{+\infty} f(x)\,\mathrm{d}x = 1$, so the total area
under the curve is 1.

Suppose we pick a point under the curve, uniformly at random, and then
look at its x-coordinate $X$. What is the PDF of $X$?

\[picture\]

Well, the probability that $X$ lies in the interval $[a,b]$ is the
probability the point we pick lies in the part of the curve between $a$
and $b$, which is the area of that area divided by the total area under
the curve 1. So
$$ \mathbb P(a \leq X \leq b) = \int_a^b f(x) \,\mathrm{d}x .$$ But
that's just the probability associated with the PDF $f$, so $X$ has PDF
$f$.

\[picture\]

This gives us a way -- in theory, at least -- to sample from a PDF $f$.
Simply pick a point at random under the curve, and take its
x-coordinate. Unfortunately, there is no known way to do this directly
in general, so it doesn't really help.

However, suppose we could find a function $h$ that (a) is positive
everywhere, (b) has finite area under the curve, (c is above $f$
everywhere, (d) and that we *could* sample a point under. We could call
such a curve an **envelope** for $f$. We could then propose a point
picked at random under the envelope $h$, reject it if it were above the
curve $f$, but accept it if it were under the curve $f$. An accepted
point would be uniform under the curve $f$, so its x-coordinate would
have PDF $f$, as required.

\[picture\]

But how can we find such an envelope $h$ we can use?

If the area under the envelope $h$ is $c$, with $1 \leq c < \infty$ (the
area must be at least 1, if the curve is above $f$), then $h$ must be
over the form $h(x) = cg(x)$ for a PDF $g$. But picking the x-coordinate
point at random under $g$ is just sampling from the random variable that
has PDF $g$. And the probability that a point with x-coordinate $x$ from
under the envelope $cg(x)$ also lies under $f$ is $f(x)/cg(x)$, so that
would be our acceptance probability.

This tells us how to perform envelope rejection sampling.

## The envelope rejection sampling algorithm

If I lost you during my motivational discussion, now is the time to
switch back on! We are going to set out the steps of the **envelope
rejection sampling** algorithm.

::: thpart
-   **Goal:** To sample from a random variable $X$ with PDF $f$.

-   **We will need:** A random variable $Y$ with PDF $g$ that we can
    sample from, and a finite constant $c$ such that $f(x) \leq cg(x)$
    for all $x$.

-   **Step 1:** Sample a proposal $Y$ from the PDF $g$.

-   **Step 2:** Accept the proposal $Y$ with probability $f(Y) / cg(Y)$;
    otherwise reject $Y$. If accepted, *end*. If rejected, *go back to
    Step 1*.
:::

Note that for given $f$ and $g$, there's no guarantee a finite constant
$c$ exists such that $f(x) \leq cg(x)$. Informally, we need $g$ to have
tails that are "at least as heavy" as the tails of $f$.

We should check this really does sample from $f$. We saw last time that
the PDF of a sample had PDF $f(x) \propto g(x) \,\alpha(x)$. Here, that
is $$ f(x) \propto g(x)\,\frac{f(x)}{cg(x)} = \frac1c\,f(x) , $$ as
required.

In envelope rejection sampling, we reject proposals $x$ with probability
$f(x)/cg(x)$. But we saw last time that rejecting samples is bad --
higher rejection probabilities leave us with fewer accepted samples (or
requiring more proposals to get the same number of accepted samples).
So, for given $f$ and $g$, it would be good to pick $c$ as small as
possible. We must have $f(x) \leq cg(x)$ for all $x$, so we must have
$c \geq f(x) / g(x)$ for all $x$, and therefore
$c \geq \sup_x f(x)/g(x)$. (Here, "sup"
means "supremum" -- it's like the maximum, except allows for the fact the
maximum might only be achieved in a limit, such as $x \to -\infty$ or
$x \to +\infty$.) The best possible $c$, therefore is to take $c$ to *equal* this maximum, $c = \sup_x f(x) / g(x)$, if it's possible to calculate it.

[**Note:** *I messed this up in the lecture, and accidentally said minimum/infimum, where it should be maximum/supremum.*]

## Examples

::: {#exm-env1}
Consider the Wigner semi-circle distribution with PDF
$$ f(x) = \frac{2}{\pi}\sqrt{1 - x^2} \qquad -1 \leq x \leq 1 . $$

We need to choose the envelope that surrounds the PDF $f$. Here's one
choice. Let $Y \sim \operatorname{U}[-1,1]$, which has PDF
$$ g(x) = \tfrac12 \qquad -1 \leq x \leq 1 ,$$ which well give a simple "box" around $f$ for the envelope. We can also generate these samples
easily, for example by using the inverse transform $Y = 2U - 1$ for
$U \sim \operatorname{U}[0,1]$.

The maximum point of $f$ is at $x = 0$, where $f(0) = \frac{2}{\pi}$ and $g(0) = \tfrac12$. So
we see that $c \geq \frac{4}{\pi}$ will be sufficient to ensure
$f(x) \leq cg(x)$ for all $x$, with the box surrounding the semi-circle. We want to take the smallest possible
$c$, so will take $c = \frac{4}{\pi}$, with the box *just* touching the semi-circle at its top.

```{r}
pdf <- function(x) (2 / pi) * sqrt(1 - x^2)
env <- function(x) (4 / pi) * (1 / 2) * (-1 <= x & x <= 1)
```

```{r semi1}
#| code-fold: TRUE
curve(
  pdf, n = 1001, from = -1, to = 1,
  xlim = c(-1.5, 1.5), ylim = c(0, 0.8), lwd = 3, col = "blue"
)
curve(env, n = 1001, add = TRUE, lwd = 2, col = "red")
```

In my code, I've set it up so that `env()` (the envelope function) is $c$ times $g$.

```{r}
n_prop <- 1e6
props <- 2 * runif(n_prop) - 1
accepts <- props[runif(n_prop) <= pdf(props) / env(props)]

length(accepts)
hist(accepts, probability = TRUE, breaks = 40)
curve(pdf, add = TRUE, lwd = 2, col = "blue")
```

We see that we accepted about 78% or 79% of proposals, and the histogram
is an excellent fit to the semicircle distribution.
:::

::: {#exm-env-norm}
Suppose we want to sample from the "half-normal" distribution
$$ f(x) = \sqrt\frac{2}{\pi} \exp\big(\tfrac12 x^2\big) \qquad x \geq 0 . $$
Let's take $Y \sim \operatorname{Exp}(1)$ with $g(x) = \mathrm{e}^{-x}$.
We know this can be easily sampled from, using the inverse transform
$Y = - \log U$. The exponential also has fatter tails than the normal,
so this should work.

To find an appropriate $c$, we consider
$$ \frac{f(x)}{g(x)} = \frac{\sqrt\frac{2}{\pi} \exp(\tfrac12 x^2)}{\exp(-x)} = \sqrt\frac{2}{\pi} \exp \big(-\tfrac12 x^2 + x\big) . $$
This is maximised where $-\tfrac12 x^2 + x$ is maximised which (by
differentiating and setting equal to 0) is at $x = 1$. So we take the
best possible $c$, which is
$$ c = \frac{f(1)}{g(1)} = \sqrt\frac{2}{\pi} \exp\big(-\tfrac12 + 1\big) = \sqrt\frac{2\mathrm{e}}{\pi} . $$

```{r}
pdf <- function(x) sqrt(2 / pi) * exp(-x^2 / 2)
env <- function(x) sqrt(2 * exp(1) / pi) * exp(-x)
```

```{r halfnorm}
#| code-fold: TRUE
curve(
  pdf, n = 1001, from = 0, to = 4,
  xlim = c(0, 3), ylim = c(0, 1.32), lwd = 3, col = "blue"
)
curve(env, n = 1001, from = 0, to = 4, add = TRUE, lwd = 2, col = "red")
```

```{r}
n_prop <- 1e6
props <- -log(runif(n_prop))
accepts <- props[runif(n_prop) <= pdf(props) / env(props)]

length(accepts)
hist(accepts, probability = TRUE, breaks = 40)
curve(pdf, add = TRUE, lwd = 2, col = "blue")
```

We accept roughly 760,000 proposals, and get an excellent fit to the
half-normal distribution.
:::

**Next time.** *We study envelope rejection sampling more closely, and
complete this part of the module on random number generation*

::: mysummary
**Summary:**

-   For envelope rejection sampling from a PDF $f$, we need a PDF $g$
    from which we can sample and a constant $c$ such that
    $f(x) \leq cg(x)$ for all $x$.

-   We propose samples from $g$, and accept a proposal $x$ with
    probability $f(x) / cg(x)$.

-   To keep the acceptance probability high, we want to pick $c$ as
    small as possible.

**Read more:** [Voss, *An Introduction to Statistical
Computing*](https://leeds.primo.exlibrisgroup.com/permalink/44LEE_INST/1fj430b/cdi_askewsholts_vlebooks_9781118728031),
Subsections 1.4.2.
:::
