Let's suppose we've forgotten the expectation of the exponential distribution $X \sim \operatorname{Exp}(2)$ with rate 2. In this simple case, we could work out the answer using the PDF $f(x) = 2\mathrm{e}^{-2x}$ as\
$$ \mathbb E X = \int_0^\infty x\,2\mathrm{e}^{-2x}\,\mathrm{d}x $$(and, without too much difficulty, get the answer $\frac12$). But instead, let's do this the Monte Carlo way.

In R, we can use the `rexp()` function to get IID samples from the exponential distribution: the full syntax is `rexp(n, rate)`, which gives `n` samples from an exponential distribution with rate `rate`. So our code here should be

```{r}
n <- 100
samples <- rexp(n, 2)
MCest <- (1 / n) * sum(samples)
MCest
```

So our Monte Carlo estimate is `{r} format(round(MCest, 5), nsmall = 5)`, to 5 decimal places.

To get a (hopefully) more accurate estimation, we can use more samples. We could also simplify the third line of this code by using the `mean()` function.

```{r}
n <- 1e6
samples <- rexp(n, 2)
MCest <- mean(samples)
MCest
```

(In the second line, `1e6` is R code for the scientific notation $1 \times 10^6$, or a million. I just picked this as "a big number, but where my code still only took a few seconds to run.")

Our new Monte Carlo estimate is `{r} format(round(MCest, 5), nsmall = 5)`, which is much closer to the true value of $\frac12$.
