Suppose we want to approximate the integral $$ \int_0^2 x^{1.6} (2-x)^{0.7} \, \mathrm{d}x . $$

Let's pick $X$ to be uniform on $[0,2]$. This means we should take $$\phi(x) = (2-0)h(x) = 2\,x^{1.6}(1-x)^{0.7}.$$ We can then approximate this integral in R using the Monte Carlo estimator $$ \int_0^2 x^{1.6} (2-x)^{0.7} \, \mathrm{d}x = \operatorname{\mathbb{E}} \phi(X) \approx \frac{1}{n} \sum_{i=1}^n 2\,X_i^{1.6} (2-X_i)^{0.7} . $$

```{r}
n <- 1e6
integrand <- function(x) x^1.6 * (2 - x)^0.7
a <- 0
b <- 2
samples <- runif(n, a, b)
mean((b - a) * integrand(samples))
```

You have perhaps noticed that, here and elsewhere, I tend to split my R code up into lots of small bits, perhaps slightly unnecessarily. After all, those 6 lines of code could simply have been written as just 2 lines

```{r}
#| eval = FALSE
samples <- runif(1e6, 0, 2)
mean(2 * samples^1.6 * (2 - samples)^0.7)
```

There's nothing *wrong* with that. However, I find that code is easier to read if divided into small pieces. It also makes it easier to tinker with, if I want to use it to solve some similar but slightly different problem.