Suppose we want to approximate the integral $$ \int_{-\infty}^{+\infty}
\mathrm{e}^{-0.1|x|} \cos x \, \mathrm{d}x . $$ This one is an integral on the whole real line, so we can't take a uniform distribution. Maybe we should take $f(x)$ to be the PDF of a normal distribution, and then put $$ \phi(x) = \frac{h(x)}{f(x)} = \frac{\mathrm{e}^{-0.1|x|} \cos x}{f(x)} . $$

But which normal distribution should we take? Well, we're *allowed* to take any one. But we'll probably get the best results if we pick one that is mostly likely to take values where $h(x)$ is big -- or, rather, where the absolute value $|h(x)|$ is big, to be precise. That is because we don't want to "waste" too many samples where $h(x)$ is very small, because they don't contribute much to the integral. But we don't want to "miss" -- or only sample very rarely -- places where $h(x)$ is big, which contribute a lot to the integral. Let's have a look at the graph of $h(x) = \mathrm{e}^{-0.1|x|} \cos x$.

```{r h-graph}
#| code-fold: true
integrand <- function(x) exp(-0.1 * abs(x)) * cos(x)

curve(
  integrand, n = 1001, from = -55, to = 55,
  col = "blue", lwd = 3,
  xlab = "x", ylab = "h(x)", xlim = c(-50,50)
)
abline(h = 0)
```

This suggests to me that a mean of 0 and a standard deviation of 20 might work quite well.

```{r}
n <- 1e6
integrand <- function(x) exp(-0.1 * abs(x)) * cos(x)
pdf       <- function(x) dnorm(x, 0, 20)
phi       <- function(x) integrand(x) / pdf(x)

samples <- rnorm(n, 0, 20)
mean(phi(samples))
```
