Let's go back to the very first example in the module, @exm-MCexp, where we were trying to find the expectation of an $\operatorname{Exp}(2)$ random variable. We used this R code:

```{r}
n <- 1e6
samples <- rexp(n, 2)
MCest <- mean(samples)
MCest
```

(Because Monte Carlo estimation is random, this won't be the *exact* same estimate we had before, of course.)

So if we want to investigate the error, we can use the sample variance of these samples. We will use the sample variance function `var()` to calculate the sample variance.

```{r}
var_est <- var(samples)
MSEest  <- var_est / n
RMSEest <- sqrt(MSEest)
c(var_est, MSEest, RMSEest)
```

The first number is `var_est` $= `r signif(var_est, 3)`$, the sample variance of our $\phi(x_i)$s: $$ s^2 = \frac{1}{n-1} \sum_{i=1}^n \big(\phi(x_i) - \widehat{\theta}_n^{\mathrm{MC}}\big)^2 . $$ This should be a good estimate of the true variance $\operatorname{Var}(\phi(X))$.

The second number is `MSEest` $= `{r} signif(var_est/n, 3)`$, our estimate of the mean-square error. Since $\operatorname{MSE}(\widehat{\theta}_n^{\mathrm{MC}}) = \frac{1}{n} \operatorname{Var}(\phi(X))$, then $\frac{1}{n} s^2$ should be a good estimate of the MSE.

The third number is `RMSEest` $= `{r} signif(sqrt(var_est/n), 3)`$ our estimate of the root-mean square error, which is simply the square-root of our estimate of the mean-square error.
