Let $X \sim \operatorname{N}(0,1)$ be a standard normal. Suppose we want to estimate $\mathbb P(X > 4)$.
We could do this the standard Monte Carlo way by sampling from $X$ itself.
$$ \widehat{\theta}_n^{\mathrm{MS}} = \frac{1}{n} \sum_{i=1}^n \mathbb I_{[4,\infty)}(X_i) . $$

However, this will not be a good estimator. To see the problem, lets run this with $n = 100\,000 = 10^5$ samples, but do it 10 times, and see what all the estimates are.

```{r}
n <- 1e5
MCest <- rep(0, 10)
for (i in 1:10) MCest[i] <- mean(rnorm(n) > 4)
MCest
```

We see a big range of values. I get different results each time I run it, but anything between $1 \times 10^{-5}$ and $8 \times 10^{-5}$, and even $0$, comes out fairly regularly as the estimate. The problem is that $X > 4$ is a very rare event -- it only comes out a handful of times (perhaps 0 to 8) out of the 100,000 samples. This means our estimate is (on average) quite inaccurate.

It would be better not to sample from $X$, but rather to sample from a distribution that is greater than 4 a more reasonably proportion of the time. Let's stick with a normal distribution with standard deviation 1, but try increasing the mean. Let's try importance sampling $Y \sim \operatorname{N}(4,1)$.

The PDFs of $X \sim \operatorname{N}(0,1)$ and $Y \sim \operatorname{N}(4,1)$ are
$$f(x) = \frac{1}{\sqrt{2\pi}} \exp\big(-\tfrac12 x^2\big) \qquad g(y) = \frac{1}{\sqrt{2\pi}} \exp\big(-\tfrac12 (y-4)^2\big) , $$
so the relevant weighting of a sample $y$ is
$$ \frac{f(y)}{g(y)} = \frac{\exp\big(-\tfrac12 y^2\big)}{\exp\big(-\tfrac12 (y-4)^2\big)} = \exp \big(\tfrac12\big(-y^2 + (y-4)^2\big)\big) = \exp(-4y+8) .  $$
So our importance sampling estimate will be
$$ \widehat{\theta}_n^{\mathrm{IS}} = \frac{1}{n} \sum_{i=1}^n \mathrm{e}^{-4y +8} \, \mathbb I_{[4,\infty)}(Y_i) . $$

Let's try this in R. Although we could use the function $\mathrm{e}^{-4y+8}$, I'll do this by using the ratios of the PDFs directly in R (just in case I made a mistake...).

```{r}
n <- 1e5
pdf_x <- function(y) dnorm(y, 0, 1)
pdf_y <- function(y) dnorm(y, 0, 4)
samples_y <- rnorm(n, 0, 4)
ISest <- mean((pdf_x(samples_y) / pdf_y(samples_y)) * (samples_y > 4))
ISest
```